---
title: "Geog4/6300: Lab 0--Tornadoes and social vulnerability"
format: gfm
execute:
  echo: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

*Your name: Mary Elizabeth Rudd*

This "lab" assignment provides an opportunity to learn the basic mechanics of Github classroom, loading data, and doing some basic data manipulation using the tools in the tidyverse. Through this lab, you will identify counties with the highest levels of social vulnerability that also had high numbers of severe tornadoes (rated F3 and higher) during this study period.

This lab assesses the following learning standards:

1. Classify variables by measurement type (nominal, ordinal, interval, ratio) and justify those classifications with reference to the characteristics of the data. (Task 2)
3. Load tabular and spatial data into a code-based environment from multiple sources, including local files, remote APIs, and spatial file formats. (Task 1)
7. Filter, aggregate, and transform datasets using grouping and summary operations to answer specific analytical questions. (Task 3, 4, and 5)
8. Join multiple datasets using appropriate join strategies and explain how different join types affect the resulting output. (Task 6 and 7)
32. Use Git and Github to create and share project materials in a repository format. (Lab submission)

## Loading the data

We'll be combining two datasets for this assignment: NOAA's database of tornadoes from 1950-2026 (so far) and the CDC's Social Vulnerability Index (SVI): https://www.atsdr.cdc.gov/place-health/php/svi/index.html

You can load the tornado data (`noaa_stormevent_tornado_2026_08.csv` in the folder `data/stormevents/`) using the `read_csv` function. The SVI data is located at `data/SVI_2022_US_county.csv`. Load the tornado dataset into an object called `tornadoes` and the SVI dataset into an object called `svi`.

(Side note: you can see how we downloaded this data directly from NWS in the "stormdata_download" script in the data folder.)

**Task 1:** *Load the tornado data and the SVI data using `read_csv`.*

```{r}
#| label: Task1

Tornado<-read.csv("data/stormevents/noaa_stormevent_tornado_2026_08.csv")
SVI<-read.csv("data/SVI_2022_US_county.csv")
```

**Task 2:** *Pick two variables from either dataset that use different levels of measurement (nominal, ordinal, interval or ratio). Explain which level best describes each variable and why. Also identify one variable that would be more efficiently read as a factor, and explain why.*

From the Tornado Dataset: 
1. Variable "State" is nominal data, because it is made of characters/qualitative data that doesn't have an inherent order
2. Variable "Injuries_Direct" is ratio data, because it is quantitative (made of numbers) and has a meaningful 0. (0 injuries is 0 injuries)
Factor efficient variable: "Month_Name" is a variable that could be efficiently read as a factor, as the months of the year (qualitative) can be sorted from January to December

## Filtering and/or summarizing the data

Open up the tornado data frame so you can look at it. There's multiple variables here related to the timing of the tornado, its location, its magnitude, and the number of injuries and fatalities. You can open the "Storm-Data-Bulk-csv_Format" pdf in the data folder to learn more about them.

For this part of the lab, you want to count the number of tornadoes rated F3 or higher (the `TOR_F_SCALE` variable) within each county (the `cty_fips` and `CZ_NAME` variables). Your resulting data frame should have one row per county and state combination, with a variable that stores the total number of severe tornadoes.

To do so, you'll need to:

1. Filter the `TOR_F_SCALE` variable to only include observations rated F3, F4, or F5.
2. Use `group_by()` to group the data by the county and state fips codes and names (`STATE_FIPS`, `STATE`, `cty_fips`, and `CZ_NAME`).
3. Use `summarise()` to count the number of these severe tornadoes within these groups.

**Task 3:** *Filter the storm data to include tornadoes rated F3 or higher and then count the number of events by county and state.*

```{r}
#| label: Task3

#insert code here
```

You also want to filter the social vulnerability data so that it only includes counties with a population over 25,000 people. This allows us to only track counties with a moderate or large population. Use the SVI Documentation pdf in the data folder to determine which variable this is, recognizing that you want the **estimate** and not the margin of error (MOE).

Filter the data below.

**Task 4:** *Filter the SVI data to include counties with a population of 25,000 or more.*

```{r}
#| label: Task4

# Your code goes here.
```

Let's add one more filter based on social vulnerability. When you look at the documentation there are a number of "dummy" variables (0/1) that flag counties that are at or above the 90th percentile for a number of factors. For example, `F_DISABL` flags counties in the top 10% for percentage of persons with a disability. There are also cumulative flags: `F_THEME2` sums all the flagged variables within the "Household Characteristics" category.

Pick one of those flags. It could be for a specific variable or a cumulative one. Your selected flag should have a plausible connection to tornado preparedness, evacuation, sheltering, recovery, or access to warnings. 

Filter the dataset you created in Task 4 so that it only includes counties you are considering vulnerable. Then list that flagged variable and explain how you chose it and (if the flag is cumulative) how you picked a numeric cutoff.

**Task 5:** *Filter the data from Task 4 to only vulnerable counties and explain your decision.*

```{r}
#| label: Task5

# Your code goes here.
```

Describe your variable, why you chose it, and (if applicable) why you chose the cutoff you did.

## Connecting the data

Next you'll need to join these tornado counts to SVI population data using the `inner_join()` function. To do so you'll need to have two fields with the same name in each dataset. The FIPS codes are in both datasets, but they have different names (`FIPS` and `cty_fips`). The following code creates a new variable called `cty_fips` in the SVI data using `rename()`.

**NOTE:** You may need to change the name of the population data below to match the object you used for the SVI data.

```{r}
#svi <- svi %>%
 # rename(cty_fips = FIPS)
```

Call the function above to rename your data. Now you're ready to join the data.

**Task 6:** *Use `inner_join()` to connect the filtered tornado data to your county data from Task 5.*

```{r}
#| label: Task6

# Your code goes here.
```

**Task 7:** *Find some documentation on the `inner_join()` function online or using help in R. Describe how it works, and explain how the results would have been different if you used `full_join()` instead.*

{Write your response here.}

## Answering your research question

**Task 8:** *Open up your joined dataset. Which three vulnerable counties have the highest number of severe tornadoes and how many were in each? Are there cities or other notable geographic features located in these counties?*

{Write your response here.}

## Challenge question

Find a peer-reviewed article published in the last ten years that uses this NWS/NOAA tornado dataset. What research question was it trying to answer? What methods were used? What are the most notable findings? Give the full citation and a summary of at least 100 words below.

Write your response here.

## Final submission stuff

**Disclosure of assistance:** Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?

Write your response here.

**Lab reflection:** How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What are the biggest things you learned by completing it?

Write your response here.

That's it! When you're done with this lab, use the Render command in Quarto to create a GitHub markdown document. Then push it to GitHub using the procedure outlined in this week's videos.