---
title: 'Geog4/6300: Lab 0'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
```

*Your name: {Your name here}*

####Change to group_by and summarise with storm data. Then join with SOVI data? Or just population data to find highest at-risk county with a pop > 250K?

This "lab" assignment provides an opportunity to learn the basic mechanics of Github classroom, loading data, and doing some basic data manipulation using the tools in the tidyverse. Through this lab, you'll identify the large county (>500K in population) with the highest number of severe tornadoes (rated F3 and higher) during this study period.

For this lab to be marked complete, the following criteria must be met:

* Your responses should show that you know how to load data from local files and remote APIs. (Task 1)
* Your responses should demonstrate that you can use basic data wrangling functions to filter and transform data (Tasks 2 and 3)
* Your responses should show that you can connect datasets using joins and understand the differences between join types (inner, left, right, full, etc.) (Tasks 4 and 5)
* Your responses should show your ability to communicate the results of yoru analysis (Task 6)

## Loading the data
We'll be combining two datasets for this assignment: NOAA's database of tornadoes from 1950-2025 (so far) and American Community Survey data on county population. You can load the tornado data (`noaa_stormevent_tornado_2025_05.csv` in the data folder) using the `read_csv` function. 

(Side note: you can see how we downloaded this data directly from NWS in the "stormdata_download" script in the data folder.)

For census data, you will use the `get_acs` function in the `tidycensus` package, which we've discussed in class. This downloads data directly from the Census via their API. For total population, you want the "B01001_001" variable and you would use "county" as the geography and "2023" as the year. Load both datasets below.

**Task 1:** _Load the tornado data and the county population data using `read_csv` and `get_acs` respectively._

```{r Task1}
#Your code goes here.
```

### Filtering and/or summarizing the data

Open up the tornado data frame so you can look at it. There's multiple variables here related to the timing of the tornado, its location, its magnitude, and the number of injuries and fatalities. You can open the "Storm-Data-Bulk-csv_Format" pdf in the data folder to learn more about them.

For this part of the lab, you want to count the number of tornadoes rated F3 or higher (the `TOR_F_SCALE` variable) within each county (the `cty_fips` and `CZ_NAME` variables). 

To do so, you'll first filter the `TOR_F_SCALE` variable to only include observations rated F3, F4, or F5. You'll then need to use `group_by` and `summarise` to count the number of these severe tornadoes by county. Do that below.

**Task 2:** _Filter the storm data to include tornadoes rated F3 or higher and then count the number of events by county._

```{r Task2}
#Your code goes here.
```

You also want to filter your county population data so that it only includes counties with a population over 500,000 people. That information is in the "estimate" variable. Do that below.

**Task 3:** _Filter the census data to include counties with a population of 500,000 or more._

```{r Task3}
#Your code goes here.
```

## Connecting the data

Next you'll need to join these tornado counts to county population using the inner_join function. To do so you'll need to have two fields with the same name in each dataset. The fips codes are in both datasets, but they have different names (`GEOID` and `cty_fips`). The following code creates a new variable called fips in the ACS data that's a character using the mutate function. 

**NOTE: You many need to change the name of the population data below to match the one you used for your county data.** 

```{r}
cty_pop<-cty_pop %>%
  rename(cty_fips=GEOID)
```

Call the function above to rename your data. Now you're ready to join the data.

**Task 4:** _Use `inner_join` to connect the filtered tornado data to the list of counties with a population of 500,000 or more_

```{r Task4}

```

**Task 5:** _Find some documentation on the inner_join function online or using help in R. Describe how it works, and explain how the results would have been different if you used full_join instead._

{Write your response here.}

### Answering your research question

### Task 6:** _Open up your joined dataset. Which three counties have the highest number of severe tornadoes and how many were in each? What cities are located within these three counties?_

{Your response goes here.}

### Challenge question
Find a peer-reviewed article published in the last ten years that uses this NWS/NOAA tornado dataset. What research question was it trying to answer? What methods were used? What are the most notable findings? Give the full citation and a summary of at least 100 words below.

{Your response goes here.}


### Final submission stuff

**Disclosure of assistance:** Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?

{Response here.}

**Lab reflection:** How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What are the biggest things you learned by completing it?

{Your response here}

That's it! When you're done with this lab, use the Knit command to render it as a Github markdown document. Then push it up to Github using the procedure outlined in this week's videos.